A Model-Driven Optimization Approach\\to Solving Partial Differential Equations

### 2014-02-12 paper/bare_conf.tex
Start writing the introduction

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{A Compiler-Driven Optimization Approach\\to Solving Partial Differential Equations}

@@

\section{Introduction}
What we do in computational science. Applications motivating our study. The programming model of lizst and op2: ``simple'' kernels applied to sets of mesh elements. The need for optimizing such kernels. Contributions of the paper:
\begin{itemize}
\item A \textbf{compiler-driven} optimisation approach that can be generalised to a wide variety of kernels employed in many computational science codes. The key feature common to these kernels is that they usually fit the L1 cache of commodity processors.
\item Automation of such code optimizations in a framework for the resoloution of partial differential equations through the Finite Element Method, called Firedrake. This allows us to perform an in-depth performance evaluation of the proposed optimization strategy in real-world computations.
\end{itemize}

\section{Background}
Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.

\section{A Compiler-driven Approach to Code Optimization}
All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead. 
Follows a description of code transformations using our compiler-driven approach.

\section{The PyOP2 Compiler}
Design and structure of the PyOP2 Compiler. Show the steps through which the IR is transformed (need to cite Firedrake here). Briefly describe a simple cost model that allows us to prune the space of code transformations.

\section{Performance Evaluation}
Performance evaluation. We show all of the results got so far and we explain them. We demonstrate the claim that a compiler-driven approach is necessary to obtain maximum performance. Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?

\section{Related Work}
Show other work of people working on optimizations for computational science kernels. Inter-kernel vectorisation paper from Istvan. Previous work in FFC. Spencer et-al + Shin et. al. attempt to optimize computations that could benefit from using BLAS, but in practice they don't, due to the very small dgemm operations employed. Saday's model-driven SIMD code vectorisation for the tensor contraction engine.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}

\bibitem{IEEEhowto:kopka}
H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

\end{thebibliography}

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{A Model-Driven Optimization Approach\\to Solving Partial Differential Equations}

@@

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs are required that model phenomena. A feature common to these scientific computations is that they involve the resolution of partial differential equations (PDEs). Numerical techniques, like Finite Volume Method (FVM) and Finite Element Method (FEM), are widely employed in real-world applications to approximate solutions of PDEs. Unstructured meshes are often used to discretize the PDE domain, since their geometric flexibility allows such PDE solvers to be extrimely effective. However, they are also the main responsible for performance degradation due to the introduction of indirect accesses (e.g. \texttt{A[B[i]]}) to access mesh components. A typical problem, therefore, is that implementing a scientific application requires a great effort, other than domain knowledge, and even more time is needed to optimize their execution time.

An unstructured mesh represents a discretization of the equation domain into two- or three-dimensional cells (or ``elements''), which define the smallest area within which the solution is sought. As the number of elements can be of the order of millions, a major issue is the time required to perform the numerical method, which can be of the order of hours or days. OP2~\cite{pyop2ics} and Lizst~\cite{lizst} are Domain Specific Languages that target mesh-based computations by allowing programmers to achieve highly-efficient parallel execution. The successful porting of Hydra, a CFD industrial application devised by Rolls Royce for turbomachinery design (based on FVM, roughly 50000 lines of code and a mesh size that can be over 100 millions edges), to OP2, demonstrates the effectiveness of this framework for implementing PDEs solvers~\cite{OP2-hydra}. Both OP2 and Lizst adopt a kernel-oriented programming model: the computation semantics is expressed through self-contained functions (or ``kernels''), each function applied in parallel to all items in a specific set of mesh components (e.g. edges, vertices, elements), with an implicit synchronization between the application of two different functions. Together with the indirections problem, having systematic optimizations for the kernels is the major concern in unstructured mesh applications.

In this paper, we describe a model-driven optimization strategy for a class of kernels used in Finite Element computations. The working set of these kernels is usually small enough to fit the L1 cache, and definitely fits the L2 cache when high-order methods are employed to improve the accuracy of the solution. However, the latter case is not considered in our work. This implies that main actors of our study will be register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.

%the focus of optimization should be on managing registers, exploiting ILP in its various forms, and reducing loop overhead

% systematic optimization strategy -> discuss the importance of domain knowledge

% A python implementation of OP2, namely PyOP2, is used in the Firedrake project as an intermediate representation for FEM computations. In Firedrake all computational kernels are generated by a compiler, (the Fenics Form Compiler, FFC), whereas in OP2 applications kernels are provided by users.

%What we do in computational science. Applications motivating our study. The programming model of lizst and op2: ``simple'' kernels applied to sets of mesh elements. The need for optimizing such kernels. Contributions of the paper:
%\begin{itemize}
%\item A \textbf{compiler-driven} optimisation approach that can be generalised to a wide variety of kernels employed in many computational science codes. The key feature common to these kernels is that they usually fit the L1 cache of commodity processors.
%\item Automation of such code optimizations in a framework for the resoloution of partial differential equations through the Finite Element Method, called Firedrake. This allows us to perform an in-depth performance evaluation of the proposed optimization strategy in real-world computations.
%\end{itemize}

\section{Background}

%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.

\section{A Model-driven Approach to Code Optimization}
%All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead. 
%Follows a description of code transformations using our compiler-driven approach.

\section{The PyOP2 Compiler}
%Design and structure of the PyOP2 Compiler. Show the steps through which the IR is transformed (need to cite Firedrake here). Briefly describe a simple cost model that allows us to prune the space of code transformations.

\section{Performance Evaluation}
%Performance evaluation. We show all of the results got so far and we explain them. We demonstrate the claim that a compiler-driven approach is necessary to obtain maximum performance. Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?

\section{Related Work}
%Show other work of people working on optimizations for computational science kernels. Inter-kernel vectorisation paper from Istvan. Previous work in FFC. Spencer et-al + Shin et. al. attempt to optimize computations that could benefit from using BLAS, but in practice they don't, due to the very small dgemm operations employed. Saday's model-driven SIMD code vectorisation for the tensor contraction engine.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
%% <OR> manually copy in the resultant .bbl file
%% set second argument of \begin to the number of references
%% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
%
%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
%
%\end{thebibliography}
%

\bibliographystyle{plain}	% (uses file "plain.bst")
\bibliography{biblio}		% expects file "myrefs.bib"
